Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-09-17 _manuscript/Manuscript.qmd
Nearly finished text updates

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy a difficult quantity to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data. Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017]. Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied or unoccupied state. In the first time step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

@@

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranging from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

DOMs have been used to study a variety of different species, though the majority of studies have been conducted on data for vertebrate taxa. They have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and DOMs have been touted as a valuable tool for modelling disease dynamics. @mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of applications have been fit for terrestrial species, though there have been a limited number of studies on aquatic systems, including @fisher2014's application to invasive salmon, @falke2012's on Great Plains stream fishes, and @pendleton2022's on whale occupancy dynamics.

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods. In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' model treating poachers and using ranger reports to create detection histories.

Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling. In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. Intepretation and drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019]. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data are collected, and how covariates responses are represented in DOMs. We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails. Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].

@@

Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate. Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable. For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. 42% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.

In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function). Statistical relationships between covariates and model parameters are represented as linear terms unless otherwise specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discussed the importance of modelling ecologically-realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates. Despite this, in our sample just 35% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships. Interactions between covariates were similarly rare, with only 26% of studies considering at least one relationship between terms. The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their typically larger range-wide scale which encompasses the full species niche, where environmental relations may be expected to be non-linear. However, as previously indicated DOMs have been frequently implemented at similarly large spatial scales where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of interactions between covariates: while they should not necessarily be used liberally, where these exist their exclusion can influence model performance [@guisan2006].

In addition to the types of covariates considered for modelling we also measured the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structure covariates which were available for use on each parameter ( @fig-covariates). This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as 1 covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates relative to initial occupancy and detection. The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant. Omission of the factors which drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles. Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs. Detection probability is also typically represented with low quantities of environmental covariates. Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple single-parameter may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested. Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22), which differ considerably in their manner of implementation. A summary of modelling practices in DOMs is given in @tbl-modelling below.

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which DOMs have been applied for. These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A. Across our study period DOMs have been frequently employed to observe trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

>>>>

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy difficult to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] . Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied (z = 1) or unoccupied state (z = 0). In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).

$$
\begin{align*}
z_1 \sim& Bernoulli(\psi_1)\\\\
z_t \sim& Bernoulli(\psi)\\\\
&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
\end{align*}
$$ {#eq-occupancy}

In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0). Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).

$$
y_{t,j} \sim Bernouilli(z_t\rho)
$$ {#eq-detection}

@@

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how the noted variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranged from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

The DOM has been used to study a variety of different species, though the majority of studies have focussed on vertebrate taxa (@fig-StudyDetails B). The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications. @mores2020 and @padilla-torres2013 have used the DOM to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon [@fisher2014], Great Plains stream fishes [@falke2012], and whales [@pendleton2022].

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxon, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as emerging methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods (@fig-StudyDetails, panel C). In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 use aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.

Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling. In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the "certain" detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. The interpretation of what 'occupancy' means and the factors which drive it may differ depending on the scale of what is considered a site, and how that relates to the ecology of a species. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. The DOM may be readily applied to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs. We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that users of DOMs have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of wetland birds. Several studies also include biotic interactions with other species as covariates, particularly those which focussing on taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].

@@

Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate (@tbl-covariates). Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These species-specific covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable. For projects interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. Forty two percent of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).

In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function). Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. Non-linear responses can be easily accommodated in the DOM by using polynomial transformations and interactions between covariates. Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates). Interactions between covariates were similarly rare, with only 26% of studies considering at least one interaction between terms (@tbl-covariates). The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear. However, as previously indicated the DOM has also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of interactions between covariates: these relationships are commonly expected based on ecological theory, and their exclusion can negatively impact model performance [@guisan2006].

In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates). This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates relative to initial occupancy and detection. The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant. Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles. Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs. Detection probability is also typically represented with low numbers of environmental covariates (@tbl-covariates). Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested. Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22) frameworks, which differ considerably in their manner of implementation. A summary of modelling practices in DOMs is given in @tbl-modelling.

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives). These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A. Across our study period DOMs have been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives). These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle. These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once. Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception. The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models. Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl. Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes. In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy. In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy. As examples, @mcgowan2020 provides projections of future occupancy for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.
